---
title: "Gaston 2"
subtitle: "A new version of gaston, with memory-mapped object objects"
author: "Juliette Meyniel and Hervé Perdry"
version: 0.1
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gaston2}
  %\VignettePackage{gaston2}
  %\VignetteDepends{gaston2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, results = "hide", message = FALSE}
oldoptions <- options()
oldpar <- par()
options(width = 85)
require(houba)
```

<style type="text/css">
h1.title {
  font-size: 200%;
}

h1 {
  font-size: 160%;
  margin: 1em 0 0.25em 0;
}

h2 {
  font-size: 130%;
  margin: 0.25em 0 0.25em 0;
}

pre {
  padding: 0px;
  margin: 0 0 0.5em 0;
}

div.sourceCode {
  margin: 0px;
}

</style>

# Gaston 2

Gaston 2 is under development. The first release is planned for mid-May 2026.

## A short overview

The R package `gaston2` is a complete rewrite of `gaston` <https://cran.r-project.org/package=gaston>.  
`Gaston` was first released in 2015 and works well for datasets of up to 10,000 individuals. `Gaston2` 
will allow users to manipulate larger datasets without loading them into memory, by using memory-mapped objects. 
It will also include additional features for dosage data.

The main features of `gaston2` will be:

* Data manipulation and filtering for quality control.
* Computation of genetic relationship matrices and principal component analysis (PCA).
* Association studies (LMM and GLMM).


## A simple example

Here is a simple example of data manipulation with `gaston2`.

```{r}
require(gaston2)
filename <- system.file("extdata", "LCT.bed", package="gaston2")

# reading in memory
x1 <- read.snp.matrix(filename)
x1
# memory mapped object
x2 <- read.snp.matrix(filename, memory = FALSE)
x2

# You can control wether the submatrices are in temporary files, or in memmory
x2[ 1:20, ]
x2[ 1:20, , "memory"]

# Computing a LD matrix as an R object
A1 <- LD(x2, c(1, ncol(x2)), measure = "r2")
str(A1)

# ...or as a memory mapped matrix, thanks to our package 'houba'
A2 <- LD(x2, c(1, ncol(x2)), measure = "r2", filename = tempfile())
A2
```

